\AorE{Wavelet Scattering Transformasies Toegepas op Walvis Vokaliserings}{Wavelet Scattering Transforms Applied to Whale Vocalisations}

### 2024-05-27 dissertation/content/conclusion.tex
draft edit nearly complete, need to do refs

Chapter \ref*{chap:p2} further expands on the ideas of chapter \ref*{chap:p1}, by including additional time-shift invariance to the \ac{tf} decomposition via first-level wavelet scattering coefficients. We additionally model the entropy as a \ac{gmm}, which is a further improvement on k-means. We lay the groundwork for efficient downsampling strategies in for wavelet scattering computations. Since wavelet scattering can be used a features, we combine the detector output with a classifier to investigate the practicality of such a system. We provide critical discussions on the dataset that is used, and highlight the real-world problems faced in evaluation \ac{pam} detections and classification systems.

@@

The papers presented in this dissertation naturally introduce wavelet scattering to both signal detection and classification for \ac{pam} systems thereby addressing some of the shortcomings of \acp{nn}. We develop a new type of scattering which is particularly well suited to the requirements of \ac{pam}.

>>>>

Chapter \ref*{chap:p2} further expands on the ideas of chapter \ref*{chap:p1}, by including additional time-shift invariance to the \ac{tf} decomposition via first-level wavelet scattering coefficients. We additionally model the entropy as a \ac{gmm}, which is a further improvement of the k-means implementation in the first paper. We lay the groundwork for efficient downsampling strategies in for wavelet scattering computations. Since wavelet scattering can be used a features, we combine the detector output with a classifier to investigate the practicality of such a system. We provide critical discussions on the dataset that is used, and highlight the real-world problems faced in evaluation \ac{pam} detections and classification systems.

@@

The papers presented in this dissertation naturally introduce wavelet scattering to both signal detection and classification for \ac{pam} systems, thereby addressing some of the shortcomings of \acp{nn}. We develop a new type of scattering which is particularly well suited to the requirements of \ac{pam}.

### 2024-05-27 dissertation/content/introduction.tex
draft edit nearly complete, need to do refs

\Ac{ml} models have become mainstay in software products and automation pipelines. Often, when there is processing power available, a \ac{ml} model will be deployed to increase efficiency, accuracy and limit human interaction. Various models, whether they are statistical, Bayesian or probabilistic, decision trees or \acfp{nn}, are configured to assist or perform various tasks across many industries.

Modern research is very focussed on advancing \ac{ml} integration, improving model performance and investigating new techniques. A large branch of this research is particularly focussed on data-driven approaches using \acp{nn}. The collection of data therefore becomes an important part of model training and deployment. \Acp{nn} present some challenges and concerns, since they require vast amounts of data, and their operation is often unpredictable and not well understood \citep{nnsurvey}. They are typically treated as ``black box'' functions, which researchers improve and design through trial and error via hyper-parameter optimisation and architecture exploration.

In the field of \ac{pam}, which focuses on monitoring wildlife using non-invasive acoustic sensors, the majority of recent \ac{sota} models are \ac{nn}-based \citep{pamsurvey}. However, metrics of real-life deployment and practical model evaluation are often ignored, since the acoustical landscape of natural environments can be surprisingly varied. For example, one cannot expect a \ac{nn} model which was trained on data that does not contain rainfall noise, to effectively operate when it starts raining. This is especially relevant for \acp{cnn}, since the learnt filters are often unexpectedly shaped in the frequency domain \cite{sincnet}.

@@

As such, designing models and systems which are robust to all the aforementioned factors are critical in automating wildlife conservation studies based on non-invasive audio sensors. This leads to the natural question of the suitability of \acp{nn} in this field and to what extent they can be used to be effective.

At the core of many audio-based systems lies a \ac{tf}-decomposition, typically \acp{mfcc} or the \ac{stft}. This dissertation focuses on \ac{tf} decompositions, and how they can be leveraged in order to reduce the usage or even entirely replace \acp{nn} in \ac{pam} systems. At the end of this investigative journey, we naturally arrive an entirely new generalisation of a feature extraction method based on \ac{ct} wavelets, which can be utilised as a drop-in replacement for a \ac{cnn} front-end. Wavelet scattering features be further improved with deep learning across filter channels \cite{nnphasecollapse}. The integration and investigation of wavelet-based processing in \ac{pam} systems are relatively scarce, even though it is very prevalent in other fields \citep{denoising,ws_fault_diag, waveletfault, waveletflight, waveletflow, waveletimagedenoising}, which we aim to remedy in the publications presented in this dissertation.

@@

In summary, some of the many challenges posed by \ac{pam} systems are:
\begin{enumerate}
    \item Long periods of silence or noise, leading to unbalanced data classes which can have a tendency for creating many false positives in models if not trained correctly.
    \item Constantly evolving noise landscape, with the interference of other bioacoustical sounds which we are often not interested in classifying.
    \item Large audio databases and recordings, which motivates the use of fast algorithms and processing techniques.
    \item Poor \ac{snr}, often less than 0 dB.
    \item Incomplete or mislabeled training data, due to poor \ac{snr} and human error.
    \item Small datasets, which only cover a portion of environment variability, making it unsuitable for a general model.
    \item Non-white, non-stationary background noise, which makes classification and signal detection non-trivial tasks.
    \item Poorly understood species behaviour in terms of their bioacoustics and movements.
\end{enumerate}

All the above factors seem demotivate the use of \acp{nn} for the following reasons: principle of operation can be unpredictable; computation time; requires reliable training data or special techniques to mitigate problematic data. Although it is possible to address these challenges, it is worth exploring non-\ac{nn} methods.

@@

Each contextual chapter is in itself an introduction to the literature and concepts presented in the following chapter. Contextual chapters serve to unify the content of the published papers and show the natural progression \ac{tf} decompositions, which finally arrives at a new wavelet scattering transform. All relevant literature is therefore gradually introduced, due to each paper conceptually expanding on the previous paper's ideas and contributions.

Although this dissertation is mostly focussed on the application to \ac{pam} systems, the theory and findings it presents are general. As such, the contextual chapters are generalised mathematical descriptions of the concepts present in each paper. Papers 1 and 2 (chapters \ref*{chap:p1}, \ref*{chap:p2}) are focussed on applying it to \ac{pam} systems. The final paper (chapter \ref*{chap:p3}) shows a generalised result from all the knowledge and investigations presented in prior papers.

@@

\begin{enumerate}
    \item \textbf{Chapter \ref*{chap:p1}: Soft-output Signal Detection for Cetacean Vocalizations using Spectral Entropy, K-means Clustering and the Continuous Wavelet Transform} -- we consider the use of the \ac{cwt} as a \ac{tf} decomposition for \ac{se} detectors, where the \ac{stft} is typically used. We then extend the conventional detector to be of a probabilistic nature. This paper introduces wavelet-based \ac{tf}decompositions in this dissertation and lays the groundwork for further extension and investigation.
    \item \textbf{Chapter \ref*{chap:p2}: Detecting and classifying Whale Calls with Wavelet Scattering and Spectral Entropy} -- we build on the work of chapter \ref*{chap:p1}, and instead consider wavelet scattering as a \ac{tf} decompistion, for which the first-level coefficients can be considered as a ``blurred'' \ac{cwt}. We further improve the \ac{se} detector using \acp{gmm}, and integrate the detector system with a classifier using wavelet scattering features. This paper focusses on the practicality of combining and improving older methods, while also introducing wavelet scattering to the field of \ac{pam}.
    \item \textbf{Chapter \ref*{chap:p3}: Separable Wavelet Scattering} -- following from the ideas presented in chapters \ref*{chap:p1} and \ref*{chap:p2}, especially with reference to computational efficiency discussed in chapter \ref*{chap:p2}, an entirely new type of wavelet scattering is generalised. We evaluate the new wavelet scattering technique with the MNIST and MedMNIST datasets for quick evaluation of the method compared to other similar wavelet scattering decompositions. We consider this paper to be the largest contribution presented in this dissertation, due to its general nature, whose design stemmed from the needs of \ac{pam} systems as discussed in section \ref*{sec:problem}.
\end{enumerate}

>>>>

\Ac{ml} models have become mainstay in software products and auto\-mation pipelines. Often, when there is processing power available, a \ac{ml} model will be deployed to increase efficiency, accuracy and limit human interaction. Various models, whether they are statistical, Bayesian or probabilistic, decision trees or \acfp{nn}, are configured to assist or perform various tasks across many industries.

Modern research is very focussed on advancing \ac{ml} integration, improving model per\-formance and investigating new techniques. A large branch of this research is particularly focussed on data-driven approaches using \acp{nn}. The collection of data therefore becomes an important part of model training and deployment. \Acp{nn} present some challenges and concerns, since they require vast amounts of data, and their operation is often unpredictable and not well understood \citep{nnsurvey}. They are typically treated as ``black box'' functions, which researchers improve and design through trial and error via hyper-para\-meter optimisation and architecture exploration.

In the field of \ac{pam}, which focuses on monitoring wildlife using non-invasive acoustic sensors, the majority of recent \ac{sota} models are \ac{nn}-based \citep{pamsurvey}. However, metrics of real-life deployment and practical model evaluation are often ignored, since the acoustical landscape of natural environments can be surpri\-singly varied. For example, one cannot expect a \ac{nn} model which was trained on data that does not contain rainfall noise, to effectively operate when it starts raining. This is especially relevant for \acp{cnn}, since the learnt filters are often unexpectedly shaped in the frequency domain \cite{sincnet}.

@@

As such, designing models and systems which are robust to all the aforementioned factors are critical in automating wildlife conservation studies with non-invasive audio sensors. This leads to the natural question of the suitability of \acp{nn} in this field and to what extent they can be used to be effective.

At the core of many audio-based systems lies a \ac{tf}-decomposition, typically \acp{mfcc} or the \ac{stft}. This dissertation focuses on \ac{tf} decompositions, and how they can be leveraged in order to reduce the usage or even entirely replace \acp{nn} in \ac{pam} systems. At the end of this investigative journey, we naturally arrive an entirely new generalisation of a feature extraction method based on \ac{ct} wavelets, which can be utilised as a drop-in replacement for a \ac{cnn} front-end. Wavelet scattering features can be further improved with deep learning across filter channels \cite{nnphasecollapse}, still making it useful in a deep learning context. The integration and investigation of wavelet-based processing in \ac{pam} systems are relatively scarce, even though it is very prevalent in other fields \citep{denoising,ws_fault_diag, waveletfault, waveletflight, waveletflow, waveletimagedenoising}, which we aim to remedy in the publications presented in this dissertation.

@@

In summary, some of the many challenges posed by \ac{pam} systems are:
\begin{enumerate}
    \item Long periods of silence or noise, leading to unbalanced data classes which can have a tendency for creating many false positives in ML models if not trained correctly.
    \item Constantly evolving noise landscape, with the interference of other unpredictable bioacoustical sounds which we are often not interested in classifying.
    \item Large audio databases and recordings, which motivates the use of fast algorithms and processing techniques.
    \item Poor \ac{snr}, often less than 0 dB.
    \item Incomplete or mislabeled training data, due to poor \ac{snr} and human error.
    \item Small datasets, which only cover a portion of environment variability, making it unsuitable for a general model.
    \item Non-white, non-stationary background noise, which can make classification and signal detection non-trivial tasks.
    \item Poorly understood species behaviour in terms of their bioacoustics and movements.
\end{enumerate}

All the above factors seem demotivate the use of \acp{nn} for the following reasons: principle of operation can be unpredictable; computation time; it requires reliable training data or special techniques to mitigate problematic data. Although it is possible to address these challenges, it is worth exploring non-\ac{nn} methods.

@@

Each contextual chapter is in itself an introduction to the literature and concepts presented in the following chapter. Contextual chapters serve to unify the content of the published papers and show the natural progression of analysing the function and purpose \ac{tf} decompositions. All relevant literature is therefore gradually introduced, since each paper conceptually expanding on the previous paper's ideas and contributions.

Although this dissertation is mostly focussed on the application to \ac{pam} systems, the theory and findings it presents are general. As such, the contextual chapters are generalised mathematical descriptions of the concepts present in each paper. Papers 1 and 2 (chapters \ref*{chap:p1}, \ref*{chap:p2}) are focussed on applying wavelet \ac{tf} decompositions to \ac{pam} systems. The final paper (chapter \ref*{chap:p3}) shows a generalised result from all the knowledge and investigations presented in prior papers.

@@

\clearpage
\begin{enumerate}
    \item \textbf{Chapter \ref*{chap:p1}: Soft-output Signal Detection for Cetacean Vocalizations using Spectral Entropy, K-means Clustering and the Continuous Wavelet Trans\-form} -- we consider the use of the \ac{cwt} as a \ac{tf} decomposition for \ac{se} detectors, where the \ac{stft} is typically used. We then extend the conventional detector to be of a probabilistic nature. This paper introduces wavelet-based \ac{tf} decompositions in this dissertation and lays the groundwork for further extension and investigation.
    \item \textbf{Chapter \ref*{chap:p2}: Detecting and classifying Whale Calls with Wavelet Scatter\-ing and Spectral Entropy} -- we build on the work of chapter \ref*{chap:p1}, and instead consider wavelet scattering as a \ac{tf} decomposition, for which the first-level coeffi\-cients can be considered as a ``blurred'' \ac{cwt}. We further improve the \ac{se} detector using \acfp{gmm}, and integrate the detector system with a classifier using wavelet scattering features. This paper focusses on the practicality of combining and improving older methods, while also introducing wavelet scattering to the field of \ac{pam}.
    \item \textbf{Chapter \ref*{chap:p3}: Separable Wavelet Scattering} -- following from the ideas presented in chapters \ref*{chap:p1} and \ref*{chap:p2}, especially with reference to computational efficiency discussed in chapter \ref*{chap:p2}, an entirely new type of wavelet scattering is generalised. We evaluate the new wavelet scattering technique with the MNIST and MedMNIST datasets to benchmark the performance of the features provided by separable scattering. We consider this paper to be the largest contribution presented in this dissertation, due to its general nature, whose design stemmed from the needs of \ac{pam} systems as discussed in section \ref*{sec:problem}.
\end{enumerate}

### 2024-05-27 dissertation/content/nomenclature.tex
added astract and conclusion
draft edit nearly complete, need to do refs

\section*{Notation}
Some notations are used interchangably, which is obvious from the context. Due to the nature of the publication format, notation is explained within the article it occurs.

@@

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    $X[k,m]$ & a time-frequency decomposition of $x$, with $k$ the frequency index and $m$ the time index \\
    $S[k,m]$ & power spectrum calculated from $X[k, m]$ \\
    $H[m]$ & spectral entropy calculated from $S[k, m]$ \\
    $X(s, b)$ & continuous wavelet transform of $x$, with $s$ as scale and $b$ as translation \\
    $\uppsi_\lambda$ & sometimes used as short-hand for  $\uppsi(\lambda t)$\\
    $\text{RM}^F_m (x)$ & rolling median filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
    $\text{RA}^F_m (x)$ & rolling average filter operator with a window of size $F$ operating on index variable $m$ on a discrete function $x$\\
    $x \triangleq y$ & $x$ is defined as $y$\\
\end{tabular}

\clearpage
\section*{Common Variables}
Some variables are used interchangably, which is obvious from the context. Due to the nature of the publication format, variables are explained within the article it occurs.

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    a & Hello \\
    a & Hello \\
    a & Hello \\
    a & Hello \\
\end{tabular}

@@

\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}

>>>>

\section*{Notation}
Some notations are used interchangably, which is obvious from the context. Due to the nature of the publication format, notation is additionally explained within the article it occurs.

@@

\section*{Notation (continued)}
\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}
    $(\vect{x})_{\downarrow \vect{r}}$ & a discrete multidimensional signal subsampled in each dimension by the components of $\vect{r}$ \\
    $X[k,m]$ & a time-frequency decomposition of a discrete signal $x$, with $k$ the frequency index and $m$ the time index \\
    $S[k,m]$ & power spectrum calculated from $X[k, m]$ \\
    $H[m]$ & spectral entropy calculated from $S[k, m]$ \\
    $X(s, b)$ & continuous wavelet transform of $x$, with $s$ as scale and $b$ as translation \\
    $\uppsi_\lambda$ & used as short-hand for a wavelet dilation -- $\lambda\uppsi(\lambda t)$\\
    $x \triangleq y$ & $x$ is defined as $y$\\
\end{tabular}

@@

\section*{Acronyms (continued)}
\begin{tabular}{p{0.15\textwidth} p{0.75\textwidth}}

### 2024-05-27 dissertation/content/p1_intro.tex
draft edit nearly complete, need to do refs

Formally, given a \ac{1d} \ac{dt} signal $x[n]$, with an index variable $n$, a \ac{tf} analysis decomposes $x$ as 
\begin{equation}
    X[k, n] = \sum_i x[n] \upsilon_k[i - n] = x * \upsilon_k,
\end{equation}
where $\upsilon_k \in \Upsilon$ is the $k$'th \ac{bpf} in the decomposition filter dictionary $\Upsilon$. Note that a \ac{lpf} can be considered to be a \ac{bpf} with a centre-frequency of 0.

Generally, the entire convolution $X[n,k]$ is not computed. Instead, only select samples are computed by speficying a hop-distance or downsample factor $d$. We indicate this by tranforming the index variable $m = dn$ as 
\begin{equation}
    \label{eqn:gentf}
    X[k, m] \triangleq X[k, dn] = \left(x * \upsilon_k\right)_{\downarrow d}.
\end{equation}

The is equivalent to applying a downsample operator $\left(\cdot\right)_{\downarrow d}$ to the convolution.

@@

A \ac{ct} Fourier transform of $x(t)$ is given as 
\begin{equation}
    \mathcal{F} \{x\} \triangleq \hat{x}(\omega) = \int_{-\infty}^{\infty} x(t) e^{-j\omega t} dt = \left(x * e^{j\omega t}\right)\bigg|_{t=0}.
\end{equation}

Derived from the conventional Fourier transform, a \ac{dt} signal $x[n], n \in \{0, ..., N-1\}$ (length $N$) has a \ac{dtft}
\begin{equation}
    \label{eqn:dtft}
    \text{DTFT}\{x\} \triangleq \hat{x}[k] = \sum_{n=0}^{N} x(t) e^{-j \frac{2\pi k n}{N}} = \left(x \otimes \upsilon_k\right)\bigg|_{n=0},
\end{equation}
for which $\upsilon_k = e^{j \frac{2\pi k n}{N}}, \ k \in \{0, ..., N-1\}$ forms an orthogonal basis of $x$. The circular convolution operator $\otimes$ implies the infinite periodic extension of $x$ beyond its boundaries of $n \in \{0, ..., N-1\}$.

Equation (\ref{eqn:dtft}) describes the \ac{dtft} as convolution evaluated at the index 0 as a useful link to the generalised notion of \ac{tf} analyses: Fourier transforms can be considered as complex sinusoidal filters being applied to the signal of interest, with periodic boundary extension.

@@

With the correct padding procedures, one can compute $x * y$ using (\ref{eqn:fftconv}).

@@

The most popular \ac{tf} analysis is the \ac{stft}. In terms of equation (\ref{eqn:gentf}), a \ac{stft} is computed with 
\begin{equation}
    \label{eqn:stftfilter}
    \upsilon_k[n] =  w[n] e^{j \frac{2\pi k n}{N}},
\end{equation}
where $w$ is a window function defined by the user. Various \acp{fft} are use to compute this transform, so the properties of $w$ is typically specified by the \ac{fft} length, and the window type (Hann, Hamming, Blackman, etc.). The hop-length (or downsample factor) $d$ is an additional parameter specified by the user. It is possible to choose $w$ and $d$ such that the \ac{stft} becomes an invertible operator -- the original signal can be entirely recovered.

@@

Suppose that we design $\Upsilon$ to contain $Q$ filters per octave. A frequency change by a factor $a$ shifts the frequency index of the fundamental to $(Q\log_2 (a) + 1) k_0$. Similarly, the $b$'th harmonic's frequency index is shifted to $(Q\log_2 (a) + 1) (Q\log_2 (b) + 1) k_0$. For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}). This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q^{-1}$ do not significantly change the frequency index structure of the \ac{tf} decomposition. The \ac{stft} is much more sensitive to such deformations across its entire structure for harmonically rich signals.

@@

$\uppsi$ can the be scaled by a factor $\lambda$ such that
\begin{equation}
    \upsilon_k(t) = \lambda_k \uppsi(\lambda_k t).
\end{equation}

@@

>>>>

Formally, given a \ac{1d} \ac{dt} signal $x[n]$, with an index variable $n$, a \ac{tf} analysis decomposes $x$ as 
\begin{equation}
    X[k, n] = \sum_i x[n] \upsilon_k[i - n] = x * \upsilon_k,
\end{equation}
where $\upsilon_k \in \Upsilon$ is the $k$'th \ac{bpf} in the decomposition filter dictionary $\Upsilon$. Note that we consider a \ac{lpf} to be a \ac{bpf} with a centre-frequency of 0.

Generally, the entire convolution $X[n,k]$ is not computed. Instead, only select samples are computed by specifying a hop-distance or downsample factor $d$. We indicate this by transforming the index variable $m = dn$ as 
\begin{equation}
    \label{eqn:gentf}
    X[k, m] \triangleq X[k, dn] = \left(x * \upsilon_k\right)_{\downarrow d}.
\end{equation}

Hop-distance is equivalently expressed by applying the downsample operator $\left(\cdot\right)_{\downarrow d}$ to the convolution.

@@

A \ac{ct} Fourier transform of $x(t)$ is given as 
\begin{equation}
    \label{eqn:fouriertransform}
    \mathcal{F} \{x\} \triangleq \hat{x}(\omega) = \int_{-\infty}^{\infty} x(t) e^{-j\omega t} dt = \left(x * e^{j\omega t}\right)\bigg|_{t=0}.
\end{equation}

Derived from the conventional Fourier transform given by equation (\ref*{eqn:fouriertransform}), a \ac{dt} signal $x[n], n \in \{0, ..., N-1\}$ of length $N$ samples has a \ac{dtft}
\begin{equation}
    \label{eqn:dtft}
    \text{DTFT}\{x\} \triangleq \hat{x}[k] = \sum_{n=0}^{N} x(t) e^{-j \frac{2\pi k n}{N}} = \left(x \otimes \upsilon_k\right)\bigg|_{n=0},
\end{equation}
for which $\upsilon_k = e^{j \frac{2\pi k n}{N}}, \ k \in \{0, ..., N-1\}$ forms an orthogonal basis of $x$. The circular convolution operator $\otimes$ implies the infinite periodic extension of $x$ beyond its boundaries of $n \in \{0, ..., N-1\}$ \citep{dspbook}.

Equation (\ref{eqn:dtft}) describes the \ac{dtft} as a convolution evaluated at the index 0 as a useful link to the generalised notion of \ac{tf} analyses: Fourier transforms can be considered as complex sinusoidal filters being applied to the signal of interest, with periodic boundary extension.

@@

With the correct padding procedures, one can compute $x * y$ using equation (\ref{eqn:fftconv}).

@@

The most popular \ac{tf} analysis is the \ac{stft}. In terms of equation (\ref{eqn:gentf}), a \ac{stft} is computed with a set of \ac{tf} filters
\begin{equation}
    \label{eqn:stftfilter}
    \upsilon_k[n] =  w[n] e^{j \frac{2\pi k n}{N}},
\end{equation}
where $w$ is a window function defined by the user. Various \acp{fft} are use to compute this transform, so the properties of $w$ is typically specified by the \ac{fft} length, and the window type (Hann, Hamming, Blackman, etc.) \citep{dspbook}. The hop-length (or downsample factor) $d$ is an additional parameter specified by the user. It is possible to choose $w$ and $d$ such that the \ac{stft} becomes an invertible operator -- the original signal can be entirely recovered \citep{stftbook}.

@@

Suppose that we design $\Upsilon$ to contain $Q$ filters per octave. A frequency change by a factor $a$ shifts the frequency index of the fundamental to $(Q\log_2 (a) + 1) k_0$. Similarly, the $b$'th harmonic's frequency index is shifted to $(Q\log_2 (a) + 1) (Q\log_2 (b) + 1) k_0$. For example, if $a=2$, the fundamental's frequency index is shifted to $k_0 + Q$ (as opposed to $2 k_0$ for the \ac{stft}). This property is referred to a frequency-shift invariance, since small frequency deformations of the order $Q^{-1}$ do not significantly change the frequency index structure of the \ac{tf} decomposition \citep{1dscattering1}. The \ac{stft} is much more sensitive to such deformations across its entire structure for harmonically rich signals.

@@

$\uppsi$ can be scaled by a factor $\lambda$ such that
\begin{equation}
    \upsilon_k(t) = \lambda_k \uppsi(\lambda_k t).
\end{equation}

@@

The multiplicate scaling, in addition to time scaling, ensures that $\uppsi_{\lambda_k}$ retains constant amplitude in the frequency domain, also referred to as L1 normalisation \cite{waveletbook}.

### 2024-05-27 dissertation/content/p2_intro.tex
added astract and conclusion
draft edit nearly complete, need to do refs

\ac{tf} decompositions are often used to extract features for signals. In audio, \acp{mfcc} are features calculated from the \ac{stft} of the signal. A ``\ac{tf}'' decomposition, as referred to in this dissertation, does not necesarrily require a time variable: spatial variables are often used. In this sense, a \ac{cnn} is can also be considered as a ``\ac{tf}'' decomposition, where multi-dimensional index variables (or index vectors) are used for ``time'' and frequency:
\begin{equation}
    X[\vect{k}, \vect{m}] = \left( x * \upsilon_{\vect{k}} \right)_{\downarrow \vect{d}}, \ \upsilon_{\vect{k}} \in \Upsilon.
\end{equation}

@@

The decomposition $X$ is often further processed by additional filters $\Phi_i[\vect{k}, \vect{m}]$, where $i$ is the index of the post-processing filter:
\begin{equation}
    \tilde{X_i}[\vect{k}, \vect{m}] = X * \Phi_i.
\end{equation}
The filters $\Phi_i$ are not necesarrily multi-dimensional, and may only operate along a single dimension of $X$. If $\Phi_i$ is an averaging (low-pass) filter, some additional stability can be observed in $\tilde{X_i}$ compared with $X$. This allows for the introduction of invariances to various deformations applied to $X$, which can be greatly beneficial for some applications. In a sense, cascased structures of $X$ and $\tilde{X_i}$ are defined in a \ac{cnn} for which the network learns the required invariances.

It is widely known that enforcing invariance in feature extraction and \ac{ml} pipelines increase performance. Enforcing invariance in \acp{nn} has been shown to improve the network, while also improving the interpretability of the network. For example, when classifying images, invariance to scale, shear, rotation and translation deformations, depending on image properties, may be useful when extracting features. Specifically, deformations of a specified order should approximately map to the same point in feature space:
\begin{equation}
    \Gamma\left(\delta\left(x\left[\vect{n}\right]\right)\right) \approx \Gamma\left(x\left[\vect{n}\right]\right),
\end{equation}
where $\Gamma$ is the feature extraction operator, and $\delta$ is the deformation operator applied to $x$.

Invariance filters ($\Phi_i$) can also have additional properties that makes it favourable for audio applications. For example, many applications ``blur'' (low-pass) the \ac{stft} magnitude or power spectrum, which can reduce the impact of noise fluctuations, and can also serve to suppress small frequency fluctuations across adjascent frequency bins (if the blurring kernel $\Phi_i$ filters across frequency as well). The blurring operation can improve the performance of signal detectors and classifiers. However, as a cost, time localisation and/or frequency localisation is reduced.

A new problem is introduced when enforcing invariances via the $\Phi_i$ filters - information is lost via averaging. Various techniques can be used to further retain information, while also retaining invariance. Wavelet scattering is such a method. In a sense, multilayer \acp{cnn} topologies with skip connections can be though of as a method to recover lost information in the deeper layers.

@@

Since \acp{mfcc} are the most widely used feature extraction method for audio signals, save for \ac{cnn} frontends which also act as a feature extractor when operating on the \ac{stft}. For this reframing, we ignore the \ac{dct} applied to the filtered values, since this linear transformation only serves to decorrelate the filter coefficients.

@@

We can view this process as a \ac{tf} decompistion with STFT filters $\upsilon_l \in \Upsilon$, which is modified with the Mel-scale triangular filters $\Phi_k$. $\upsilon_l$ provides time-shift invariance due to the STFT window, whereas $\Phi_k$ selectively averages frequency content of the STFT, thereby providing some frequency-shift invariance.

@@

The notation in equations (\ref*{eqn:general_scalogram}) and (\ref*{eqn:general_scattering}) conforms the general \ac{tf} description of chapter \ref*{chap:p1i}, although the specific wavelet scattering operators in chapters \ref*{chap:p2} to \ref*{chap:p3} follow a different standard. These operators are known as wavelet scattering, which formalise the notion of ``averaging a \ac{tf}-decomposition'' in terms of signal processing theory. Chapters \ref*{chap:p2} to \ref*{chap:p3} discuss the specifics, flavours and implementation of scattering operators at length.

>>>>

\ac{tf} decompositions are often used to extract features from signals. In audio, \acp{mfcc} are features calculated from the \ac{stft} of the signal. Note that a ``\ac{tf}'' decomposition, as referred to in this dissertation, does not necesarrily require a time variable: spatial variables are often present instead. In this sense, a \ac{cnn} is can also be considered as a ``\ac{tf}'' decomposition, where multi-dimensional index variables (or index vectors) are used for ``time'' and frequency:
\begin{equation}
    X[\vect{k}, \vect{m}] = \left( x * \upsilon_{\vect{k}} \right)_{\downarrow \vect{d}}, \ \upsilon_{\vect{k}} \in \Upsilon.
\end{equation}

@@

The decomposition $X$ is often further processed by additional filters $\Phi_i[\vect{k}, \vect{m}]$, where $i$ is the index of the post-processing filter:
\begin{equation}
    \tilde{X_i}[\vect{k}, \vect{m}] = X * \Phi_i.
\end{equation}
The filters $\Phi_i$ are not necesarrily multi-dimensional, and may only operate along a single dimension of $X$. If $\Phi_i$ is an averaging (low-pass) filter, some additional stability can be observed in $\tilde{X_i}$ compared with $X$. This allows for the introduction of invariances to various deformations applied to $X$, which can be greatly beneficial for some applications \citep{cnninvariance}. In a sense, cascased structures of $X$ and $\tilde{X_i}$ are defined in a \ac{cnn} for which the network learns the required invariances \cite{2dscattering}.

It is widely known that enforcing invariance in feature extraction and \ac{ml} pipelines increase performance. Enforcing invariance in \acp{nn} has been shown to improve the network, while also improving the interpretability of the network \cite{cnninvariance2}. For example, when classifying images, invariance to scale, shear, rotation and translation deformations, de\-pending on image properties, may be useful when extracting features. Specifically, defor\-mations of a specified order should approximately map to the same point in feature space:
\begin{equation}
    \Gamma\left(\delta\left(x\left[\vect{n}\right]\right)\right) \approx \Gamma\left(x\left[\vect{n}\right]\right),
\end{equation}
where $\Gamma$ is the feature extraction operator, and $\delta$ is the deformation operator applied to $x$.

Invariance filters ($\Phi_i$) can also have additional properties that makes it favourable for audio applications. For example, many applications ``blur'' (low-pass) the TF decomposition \citep{seyicwt, speechmodulation}, which can reduce the impact of noise fluctuations, and can also serve to suppress small frequency fluctuations across adjascent frequency bins (if the blurring kernel $\Phi_i$ filters across frequency as well). The blurring operation can improve the performance of signal detectors and classifiers. However, as a cost, time localisation and/or frequency localisation is reduced.

A new problem is introduced when enforcing invariances via the $\Phi_i$ filters - information is lost via averaging \cite{1dscattering1}. Various techniques can be used to further retain information, while also retaining invariance. Wavelet scattering is such a method \cite{ws,2dscattering}. In a sense, multilayer \acp{cnn} topologies with skip connections can be thought of as a method to recover lost information in the deeper layers.

@@

\acp{mfcc} are the most widely used feature extraction method for audio signals \citep{mfccreview}, save for \ac{cnn} frontends which also act as a feature extractor when operating on the \ac{stft}. For this reframing, we ignore the \ac{dct} applied to the filtered values, since this linear transformation only serves to decorrelate the filter coefficients \citep{dctdecorrelation}.

@@

We can view this process as a \ac{tf} decomposition with STFT filters $\upsilon_l \in \Upsilon$, which is modified with the Mel-scale triangular filters $\Phi_k$. $\upsilon_l$ provides time-shift invariance due to the STFT window, whereas $\Phi_k$ selectively averages frequency content of the STFT, thereby providing some frequency-shift invariance.

@@

The notation in equations (\ref*{eqn:general_scalogram}) and (\ref*{eqn:general_scattering}) conforms the general \ac{tf} description of chapter \ref*{chap:p1i}, although the specific wavelet scattering operators in chapters \ref*{chap:p2} to \ref*{chap:p3} follow standard notation \citep{waveletbook}. These operators are known as wavelet scattering, which formalise the notion of ``averaging a \ac{tf}-decomposition'' in terms of signal processing theory. Chapters \ref*{chap:p2} to \ref*{chap:p3} discuss the specifics, flavours and implementation of scattering operators at length.
